Explainability and Adversarial Robustness for RNNs

### 2019-12-19 paper/paper.tex
Small text changes

Network communication is usually grouped into \textit{flows}, which are commonly defined as a sequence of packets sharing certain properties. %, between two applications potentially being located at different physical locations.
When analyzing flows, not only the aforementioned features are available but also features related to the timing of the individual packets. % as well as their size. %AH: don't see why pkt size is not available without flows
%For example, a specific attack might require the attacker to send three small packets and the victim to reply with two large packets followed by a small one to the attacker.
Various approaches have been proposed to extract features from flows and then perform anomaly detection with the extracted flows \cite{meghdouri_analysis_2018}.
While these approaches frequently work well, it is problematic that the whole flow has to be received first and only afterwards anomaly detection is applied, revealing attack flows. Thus, we design a network \gls{ids} that operates on a per-packet basis and decides if a packet is anomalous based on features that are available even for traffic that is encrypted above the transport layer, like for example TLS or QUIC.
At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
We show that our system has similar performance to other flow-based anomaly detection systems but can detect anomalies \textit{before} the flow is over.

@@

Based on the insights gained by the feature importance and explainability methods, we propose two defence methods.
%develop a simple method for defending against adversaries who can manipulate arbitrary packets by simply leaving out potentially manipulable information from the classification process.
First, by simply leaving out manipulable features, we obtain a classifier which is slightly less accurate but is still useful to be deployed in a real scenario. Second,
by deploying an adversarial training procedure,
%AH. compared to other research about adversarial training, I dont feel that out approach is so special
%develop a custom adversarial training procedure which especially focuses on ease of use and low resource consumption and aims to keep the number of additionally introduced training hyperparameters low. With this novel strategy
we can reduce the attack surface and harden the resulting \gls{ids} while retaining similar classification performance. % like for the original \gls{ml} model.

@@

\subsection{Carlini-Wagner}
We implemented the \gls{cw} \cite{carlini_towards_2017}, performing gradient descent on the optimization objective
\begin{equation} \label{eq:carliniWagner}
d(X,\tilde X) + \kappa  \max(Z(\tilde X), \delta).
\end{equation}
Here, $d(\cdot)$ is a distance metric and $\kappa \in \mathbb R^+$ is a parameter governing the tradeoff achieved between attack success and distance from the original flow. Furthermore, $Z(\cdot)$ denotes the neural network's logit output, $X$ denotes the original flow and $\tilde X$ the adversarial flow under optimization.

$\delta \in \mathbb R$ is a parameter that determines how far an adversary wants to exceed the decision boundary.
In the original publication $\delta=0$, meaning that the network's decision for adversarial samples is just between attack and benign traffic. In the present context, we need to make sure that the classifier's prediction would actually be benign, even though a certain level of noise will be added to \glspl{iat} due to the network between attacker and victim. Hence, we introduced $\delta=-0.2$, corresponding to a prediction confidence of 55\% for the sample to be benign.

We used $L_1$ distance as distance metric, as we consider it to represent practically relevant differences of network flows best. We used \gls{pgd} for meeting the real-world constraints discussed above.

@@

As expected, \gls{cw} delivers the best performance and FGSM, while being the fastest of all investigated algorithms, delivers the worst performance.
%\autoref{fig:adv_per_family} depicts how successful the attack is for individual attack families.
The figure shows significant differences for detecting different attack families in the first place. Also for generating adversarial flows some attack families are more susceptible than others. However, the results match to a high degree with our expectations. For example, SQL injection attacks apparently are closer to benign traffic than \gls{dos} attacks and, hence, finding adversarial samples should be easier.

@@

\autoref{fig:recall_ars} shows that certain attack types, like ``DoS SlowHTTPTest'', are easy to classify for an IDS but still are also surprisingly vulnerable to adversarial modifications by an attacker to make them look benign.

@@

\begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
\item
\textbf{Feature quantity.}
The number of features fed into an \gls{rnn} is number of packet features times the length of the flow. For long flows, the total number of inputs can become very large.

>>>>

Network communication is usually aggregated into \textit{flows}, which are commonly defined as a sequence of packets sharing certain properties. %, between two applications potentially being located at different physical locations.
When analyzing flows, not only the aforementioned features are available but also features related to the timing of the individual packets. % as well as their size. %AH: don't see why pkt size is not available without flows
%For example, a specific attack might require the attacker to send three small packets and the victim to reply with two large packets followed by a small one to the attacker.
Various approaches have been proposed to extract features from flows and then perform anomaly detection with the extracted flows \cite{meghdouri_analysis_2018}.
While these approaches frequently work well, it is problematic that the whole flow has to be received first and only afterwards anomaly detection is applied, revealing attack flows. Thus, we design a network \gls{ids} that operates on a per-packet basis and decides if a packet is anomalous based on features that are available even for traffic that is encrypted above the transport layer, like for example TLS or QUIC.
At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
We show that our system has similar performance to other flow-based anomaly detection systems but can detect anomalies \textit{before} the flow is over.

@@

Based on the insights gained by the feature importance and explainability methods, we propose two defence methods.
%develop a simple method for defending against adversaries who can manipulate arbitrary packets by simply leaving out potentially manipulable information from the classification process.
First, by simply leaving out manipulable features, we obtain a classifier which is slightly less accurate but is still useful to be deployed in a real scenario. Second,
by deploying an adversarial training procedure,
%AH. compared to other research about adversarial training, I dont feel that out approach is so special
%develop a custom adversarial training procedure which especially focuses on ease of use and low resource consumption and aims to keep the number of additionally introduced training hyperparameters low. With this novel strategy
we can reduce the attack surface and harden the resulting \gls{ids} while retaining all features and similar classification performance. % like for the original \gls{ml} model.

@@

\subsection{Carlini-Wagner}
We implemented the \gls{cw} \cite{carlini_towards_2017}, performing gradient descent on the optimization objective
\begin{equation} \label{eq:carliniWagner}
d(X,\tilde X) + \kappa  \max(Z(\tilde X), \delta).
\end{equation}
Here, $d(\cdot)$ is a distance metric and $\kappa \in \mathbb R^+$ is a parameter governing the tradeoff achieved between attack success and distance from the original flow. Furthermore, $Z(\cdot)$ denotes the neural network's logit output, $X$ denotes the original flow and $\tilde X$ the adversarial flow optimized by \gls{cw}.

$\delta \in \mathbb R$ is a parameter that determines how far an adversary wants to exceed the decision boundary.
In the original publication $\delta=0$, meaning that the network's decision for adversarial samples is just between attack and benign traffic. In the present context, we need to make sure that the classifier's prediction would actually be benign, even though a certain level of noise will be added to \glspl{iat} due to the network between attacker and victim. Hence, we introduced $\delta=-0.2$, corresponding to a prediction confidence of 55\% for the sample to be benign after the sigmoid activation function.

We used $L_1$ as distance metric, as we consider $L_1$ distance to represent practically relevant differences of network flows best. We used \gls{pgd} for meeting the real-world constraints discussed above.

@@

%As expected,
\gls{cw} delivers the best performance and FGSM, while being the fastest of all investigated algorithms, delivers the worst performance.
%\autoref{fig:adv_per_family} depicts how successful the attack is for individual attack families.
The figure shows significant differences for detecting different attack families in the first place. Also for generating adversarial flows some attack families are more susceptible than others. However, the results match to a high degree with our expectations. For example, SQL injection attacks apparently are closer to benign traffic than \gls{dos} attacks and, hence, finding adversarial samples should be easier.

@@

\autoref{fig:recall_ars} shows that certain attack types, like ``DoS SlowHTTPTest'', are easy to classify for an IDS but still are also surprisingly vulnerable to adversarial modifications by an attacker aiming to make them look benign.

@@

\begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
\item
\textbf{Feature quantity.}
The number of features fed into an \gls{rnn} is the number of packet features times the length of the flow. For long flows, the total number of inputs can become very large.
